Handling missing data in smartphone location logs

### 2018-02-09 report/thesisReportApa.tex
reorganised folders

Within behavioural sciences, researchers have used GPS data to
investigate wideranging topics. For example, Zenk, Schulz, and
Odoms-Young (2009) investigate the effects of the food environment on
eating patterns. Harari et al. (2016) look at the movement
correlates of personality, finding that extroverted individuals spend less
time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
(2015) looks at how academic performance is affected by movement
patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
depression.

@@

Missing data is a pervasive issue as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

@@

The accuracy of smartphone location measurements is substantially lower
than that of professional GPS trackers. Android phones collect location
information through a variety of methods. Other than GPS measurements,
Androids use less-accurate heuristics such as WiFi access points and
cellphone triangulation. Different methods are used because of
computational and battery constraints. GPS is the most energy consuming
sensor on most smartphones (Chen et al., 2006; LaMarca et al.,
2005). In professional GPS trackers less than 80\% of measurements fall
within 10 meters of the true location. GPS measures are most inaccurate
in dense urban locations and indoors (Duncan et al., 2013;
Schipperijn et al., 2014). Unfortunately for researchers, this is where
people in the developed world spend most of their time.

@@

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information. For instance, Palmius et al.
(2017) calculate several measures to predict bipolar depression, such as
the total distance travelled, the amount of time spent at home,
travelling and the regularity of diurnal movement. As an example we
calculate two of these measures: time spent at home and distance
travelled for the month of February. Figure \ref{fig:aggrePlot} shows
how the aggregate measures are influenced by the noise in the
measurements and the missingness. In the case of distance travelled,
simply filtering out the most inaccurate measures leads to significantly
lower estimates. This phenomenon is mapped out on Figure
\ref{fig:accuracyPlot} . On the other hand, the estimated time spent at
home on any given day can depend greatly on the imputation method due to
the extent of missing data.

@@

In climate or geological research spatiotemporal imputation methods are
often used. For instance, the CUTOFF method estimates missing values
using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
2014 ). The authors illustrate their example using rainfall data from
gauging stations across Australia. Similarly, Zhang et al. (2017) use
a variety of machine learning methods to impute missing values. The
example provided relates to underground water data. Generally these
models assume fixed measurement stations (such as rainfall gauging
stations).

@@

Each imputation method used each of the three filtering methods as an
input. Thus in the end we have seven methods to evaluate: three
for each filtering method as well as four for each imputation method.

>>>>

Within behavioural sciences, researchers have used GPS data to
investigate wideranging topics. For example, Zenk, Schulz, and
Odoms-Young (2009) investigate the effects of the food environment on
eating patterns. G. M. Harari et al. (2016) look at the movement
correlates of personality, finding that extrovers individuals spend less
time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
(2015) looks at how academic performance is affected by movement
patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
depression.

@@

Missing data is a pervasive issue as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. G. M.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

@@

The accuracy of smartphone location measurements is substantially lower
than that of professional GPS trackers. Android phones collect location
information through a variety of methods. Other than GPS measurements,
Androids use less-accurate heuristics such as WiFi access points and
cellphone triangulation. Different methods are used because of
computational and battery constraints. GPS is the most energy consuming
sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
2005). In professional GPS trackers less than 80\% of measurements fall
within 10 meters of the true location. GPS measures are most inaccurate
in dense urban locations and indoors (S. Duncan et al., 2013;
Schipperijn et al., 2014). Unfortunately for researchers, this is where
people in the developed world spend most of their time.

@@

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information. For instance, Palmius et al.
(2017) calculate several measures to predict bipolar depression, such as
the total distance travelled, the amount of time spent at home,
travelling and the regularity of diurnal movement. As an example we
calculate two of these measures: time spent at home and distance
travelled for the month of February. Figure \ref{fig:aggrePlot} shows
how the aggregate measures are influenced by the noise in the
measurements and the missingness. In the case of distance travelled,
simply filtering out the most inaccurate measures leads to significantly
lower estimates. This phenomenon is mapped out on Figure
\ref{fig:accuracyPlot} . On the other hand, the time spent at home would
be greatly influenced at times by the imputation method, as several days
with long periods of missing data.

@@

In climate or geological research spatiotemporal imputation methods are
often used. For instance, the CUTOFF method estimates missing values
using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
2014 ). The authors illustrate their example using rainfall data from
gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
a variety of machine learning methods to impute missing values. The
example provided relates to underground water data. Generally these
models assume fixed measurement stations (such as rainfall gauging
stations).

@@

Each imputation method used each of the three filtering methods as an
input. Thus in the end we end up with eight methods to evaluate, three
for each filtering method as well as four for each imputation method.
